Hydra battles in Coq

### 2020-10-07 doc/hydras.tex
minor modifications

Schütte's definition of countable ordinal relies on the following three axioms.

\begin{enumerate}
\item There exists a well-ordered set $(\mathbb{O},<)$
\item Every bounded subset of $\mathbb{O}$  is countable
\item Every countable subset of $\mathbb{O}$  is bounded.
\end{enumerate}

@@

\begin{itemize}
\item First, please keep in mind  that the set of countable ordinals is not countable. Thus, we cannot hope to represent all countable ordinals as finite terms of an inductive type, which was possible with  the set of ordinals strictly less than $\epsilon_0$ (resp. $\Gamma_0$)
\item We tried to be as close as possible to K. Schütte's text, which uses ``classical'' mathematics : excluded middle, Hilbert's $\epsilon$ (choice) and Russel's $\iota$ (definite description) operators. Both operators allow us to write definitions close to the natural mathematical language, such as ``$\textrm{succ}$ is \emph{the} least ordinal strictly greater than $\alpha$''
\item Please note that only the sub-library \href{../src/html/hydras.Schutte.Schutte.html}{Ordinals.Schutte.Schutte/} is ``contaminated'' by axioms, and that the rest of our libraries remain constructive.

@@

\begin{Coqsrc}
Parameter Ord : Type.
Parameter lt : relation Ord.
Notation "a o< b" := (lt a b): schutte_scope.

@@

\begin{Coqsrc}
Axiom AX2 : forall X: Ensemble Ord, 
   (exists a,  (forall y, In X y -> y o< a)) ->
   countable X.

Axiom AX3 : forall X : Ensemble Ord,
              countable X -> 
              exists a,  forall y, In X y -> y o< a.
\end{Coqsrc}

@@

We are now able de define \texttt{zero} as the least ordinal. For this purpose,
we define a function returning the least element of any [non-empty]  set.

@@

\begin{Coqsrc}
Lemma zero_le (alpha : Ord) :  zero o<= alpha.
Proof.
  unfold zero, the_least, the; apply iota_ind.
\end{Coqsrc}

@@

subgoal 2 (ID 156) is:
 forall a : Ord, unique (least_member lt ordinal) a -> 
                a o<= alpha
\end{Coqanswer}

@@

Since we won't be able to prove the proposition
\linebreak \Verb|{exists! a: Ord, least_member (Empty_set Ord) a|, the only properties we would be able to prove about \texttt{bottom} would be \emph{trivial} properties, 
\emph{i.e.}, satisfied by \emph{any} element of type \texttt{Ord}, like for instance
\texttt{bottom = bottom}, or \texttt{zero o<= bottom}.

\begin{Coqbad}
Lemma le_zero_bottom : zero o<= bottom. 
Proof. apply zero_le. Qed.

@@

Lemma le_bottom_zero : bottom o<= zero.
Proof.
   unfold bottom, the_least, the; apply iota_ind.
\end{Coqbad}

@@

subgoal 2 (ID 414) is:
 forall a : Ord, unique (least_member lt (Empty_set Ord)) a -> 
      a o<= zero
\end{Coqanswer}

@@

\begin{Coqsrc}
Definition succ (alpha : Ord) := the_least (fun beta => alpha o< beta).
\end{Coqsrc}

@@

It is also easy to define recursively the finite ordinals.

@@

\begin{Coqsrc}
Definition succ_spec (alpha:Ord) :=
  least_member   lt (fun z => alpha o< z).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_succ (alpha : Ord) :  alpha o< succ alpha.
Proof.
  destruct  (succ_ok  alpha);  tauto.
Qed.

@@

Lemma lt_succ_le (alpha beta : Ord):
  alpha o< beta -> succ alpha o<= beta.
Proof with eauto with schutte.
  intros  H;  pattern (succ alpha); apply the_least_ok ... 
  exists (succ alpha); red;apply lt_succ ...
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma lt_succ_le_2 (alpha beta : Ord):
  alpha o< succ beta -> alpha o<= beta.

Lemma succ_mono (alpha beta : Ord):
  alpha o< beta -> succ alpha o< succ beta.

Lemma succ_monoR (alpha beta : Ord) :
 succ alpha o< succ beta -> alpha o< beta.

Lemma lt_succ_lt (alpha beta : Ord) :
  is_limit beta ->  alpha o< beta -> succ alpha o< beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma finite_lt_omega : forall i: nat,  i o< omega.

Lemma lt_omega_finite alpha : Ord) : 
  alpha o< omega ->  exists i:nat, alpha =  i.

@@

\begin{Coqsrc}
 Definition segment (A: Ensemble Ord) :=
  forall alpha beta, In A alpha -> beta o< alpha -> In A  beta.

@@

\begin{Coqsrc}
Definition ordering_function (f : Ord -> Ord)(A B : Ensemble Ord) :=
 segment A /\
 (forall a, In A a -> In B (f a)) /\
 (forall b, B b -> exists a, In A a /\ f a = b) /\
 forall a b, In A a -> In A b -> a o< b ->  f a o< f b.

@@

\begin{Coqsrc}
Theorem ordering_le : forall f A B,
    ordering_function f A B ->
    forall alpha, In A alpha -> alpha  o<= f alpha.

@@

Lemma le_plus_l (alpha beta : Ord) : alpha o<= alpha + beta.

Lemma le_plus_r (alpha beta : Ord) :  beta o<= alpha + beta.

Lemma plus_mono_r (alpha beta gamma : Ord) : 
    beta o< gamma -> alpha + beta o< alpha + gamma.

@@

Lemma finite_plus_infinite (n : nat) (alpha : Ord) :
  omega o<= alpha -> n + alpha = alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition AP : Ensemble Ord :=
  fun alpha => 
  zero o< alpha /\
  (forall beta, beta o< alpha ->  beta + alpha = alpha).
\end{Coqsrc}

@@

Lemma omega_second_AP :
  least_member   lt 
                  (fun alpha => 1 o< alpha /\ In AP alpha)
                  omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma AP_plus_closed (alpha beta gamma : Ord): 
     In AP alpha -> beta o< alpha -> gamma o< alpha -> beta + gamma o< alpha.

@@

Lemma phi0_mono (alpha beta : Ord) :
  alpha o< beta ->  phi0 alpha o< phi0 beta.

@@

Lemma is_limit_phi0 (alpha : Ord) :
  zero o< alpha ->  is_limit (phi0 alpha).

@@

Lemma phi0_le (alpha : Ord) : alpha o<= phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_phi0 (alpha : Ord):
  alpha o< epsilon0 -> alpha o< phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Cr_fun : forall alpha : Ord,
       (forall beta : Ord, beta o< alpha -> Ensemble Ord) ->
        Ensemble Ord 
:= 
   fun (alpha :Ord)
        (Cr : forall beta, 
                beta o< alpha -> Ensemble Ord) 
        (x : Ord) => (
       (alpha = zero /\ AP x) \/
       (zero o< alpha /\
        forall beta (H:beta o< alpha),
          the_ordering_segment (Cr beta H) x /\ ord (Cr  beta H) x = x)).

@@

Definition sorted (l: cnf_t) :=
  LocallySorted (fun alpha beta => beta o<= alpha) l.

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0 : 
 forall l alpha, 
   is_cnf_of alpha l ->  alpha o< epsilon0 ->
   Forall (fun beta =>  beta o< alpha) l.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0_iff : 
 forall l alpha, 
   is_cnf_of alpha l ->  
   (alpha o< epsilon0 <->  Forall (fun beta =>  beta o< alpha) l).
\end{Coqsrc}

@@

Theorem inject_mono (beta gamma : T1) :
  T1.lt  beta gamma -> 
  T1.nf beta -> T1.nf gamma -> 
  inject beta o< inject gamma.

@@

\begin{Coqsrc}
Theorem inject_lt_epsilon0 (alpha : T1):
      inject alpha o< epsilon0.

>>>>

Schütte's definition of countable ordinals relies on the following three axioms:

There  exists a strictly ordered set , such that
\begin{enumerate}
\item  $(\mathbb{O},<)$ is well-ordered
\item Every bounded subset of $\mathbb{O}$  is countable
\item Every countable subset of $\mathbb{O}$  is bounded.
\end{enumerate}

@@

\begin{itemize}
\item First, please keep in mind  that the set of countable ordinals is not countable. Thus, we cannot hope to represent all countable ordinals as finite terms of an inductive type, which was possible with  the set of ordinals strictly less than $\epsilon_0$ (resp. $\Gamma_0$)
\item We tried to be as close as possible to K. Schütte's text, which uses ``classical'' mathematics : excluded middle, Hilbert's $\epsilon$ (choice) and Russel's $\iota$ (definite description) operators. Both operators allow us to write definitions close to the natural mathematical language, such as ``$\textrm{succ}$ is \emph{the} least ordinal strictly greater than $\alpha$''
\item Please note that only the library \href{../src/html/hydras.Schutte.Schutte.html}{Schutte/*.v} is ``contaminated'' by axioms, and that the rest of our libraries remain constructive.
\end{itemize}

@@

\begin{Coqsrc}
Parameter Ord : Type.
Parameter lt : relation Ord.
Infix "<" := lt : schutte_scope.

@@

\begin{Coqsrc}
Axiom AX2 : forall X: Ensemble Ord, 
   (exists a,  (forall y, In X y -> y < a)) ->
   countable X.

Axiom AX3 : forall X : Ensemble Ord,
              countable X -> 
              exists a,  forall y, In X y -> y < a.
\end{Coqsrc}

@@

We are now able to define \texttt{zero} as the least ordinal. For this purpose,
we define a function returning the least element of any [non-empty]  subset.

@@

\begin{Coqsrc}
Lemma zero_le (alpha : Ord) :  zero <= alpha.
Proof.
  unfold zero, the_least, the; apply iota_ind.
\end{Coqsrc}

@@

subgoal 2 (ID 156) is:
 forall a : Ord, unique (least_member lt ordinal) a -> 
                a <= alpha
\end{Coqanswer}

@@

Since we won't be able to prove the proposition
\linebreak \Verb|{exists! a: Ord, least_member (Empty_set Ord) a|, the only properties we would be able to prove about \texttt{bottom} would be \emph{trivial} properties, 
\emph{i.e.}, satisfied by \emph{any} element of type \texttt{Ord}, like for instance
\texttt{bottom = bottom}, or \texttt{zero <= bottom}.

\begin{Coqbad}
Lemma le_zero_bottom : zero <= bottom. 
Proof. apply zero_le. Qed.

@@

Lemma le_bottom_zero : bottom <= zero.
Proof.
   unfold bottom, the_least, the; apply iota_ind.
\end{Coqbad}

@@

subgoal 2 (ID 414) is:
 forall a : Ord, unique (least_member lt (Empty_set Ord)) a -> 
      a <= zero
\end{Coqanswer}

@@

\begin{Coqsrc}
Definition succ (alpha : Ord) := the_least (fun beta => alpha < beta).
\end{Coqsrc}

@@

@@

\begin{Coqsrc}
Definition succ_spec (alpha:Ord) :=
  least_member   lt (fun z => alpha < z).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_succ (alpha : Ord) :  alpha < succ alpha.
Proof.
  destruct  (succ_ok  alpha);  tauto.
Qed.

@@

Lemma lt_succ_le (alpha beta : Ord):
  alpha < beta -> succ alpha <= beta.
Proof with eauto with schutte.
  intros  H;  pattern (succ alpha); apply the_least_ok ... 
  exists (succ alpha); red;apply lt_succ ...
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma lt_succ_le_2 (alpha beta : Ord):
  alpha < succ beta -> alpha <= beta.

Lemma succ_mono (alpha beta : Ord):
  alpha < beta -> succ alpha < succ beta.

Lemma succ_monoR (alpha beta : Ord) :
 succ alpha < succ beta -> alpha < beta.

Lemma lt_succ_lt (alpha beta : Ord) :
  is_limit beta ->  alpha < beta -> succ alpha < beta.
\end{Coqsrc}

\section{Finite ordinals}

Using \texttt{succ}, it is now easy to define recursively all the finite ordinals.

@@

\begin{Coqsrc}
Lemma finite_lt_omega : forall i: nat,  i < omega.

Lemma lt_omega_finite alpha : Ord) : 
  alpha < omega ->  exists i:nat, alpha =  i.

@@

\begin{Coqsrc}
 Definition segment (A: Ensemble Ord) :=
  forall alpha beta, In A alpha -> beta < alpha -> In A  beta.

@@

\begin{Coqsrc}
Definition ordering_function (f : Ord -> Ord)(A B : Ensemble Ord) :=
 segment A /\
 (forall a, In A a -> In B (f a)) /\
 (forall b, B b -> exists a, In A a /\ f a = b) /\
 forall a b, In A a -> In A b -> a < b ->  f a < f b.

@@

\begin{Coqsrc}
Theorem ordering_le : forall f A B,
    ordering_function f A B ->
    forall alpha, In A alpha -> alpha <= f alpha.

@@

Lemma le_plus_l (alpha beta : Ord) : alpha <= alpha + beta.

Lemma le_plus_r (alpha beta : Ord) :  beta <= alpha + beta.

Lemma plus_mono_r (alpha beta gamma : Ord) : 
    beta < gamma -> alpha + beta < alpha + gamma.

@@

Lemma finite_plus_infinite (n : nat) (alpha : Ord) :
  omega <= alpha -> n + alpha = alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition AP : Ensemble Ord :=
  fun alpha => 
  zero < alpha /\
  (forall beta, beta < alpha ->  beta + alpha = alpha).
\end{Coqsrc}

@@

Lemma omega_second_AP :
  least_member   lt 
                  (fun alpha => 1 < alpha /\ In AP alpha)
                  omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma AP_plus_closed (alpha beta gamma : Ord): 
     In AP alpha -> beta < alpha -> gamma < alpha -> beta + gamma < alpha.

@@

Lemma phi0_mono (alpha beta : Ord) :
  alpha < beta ->  phi0 alpha < phi0 beta.

@@

Lemma is_limit_phi0 (alpha : Ord) :
  zero < alpha ->  is_limit (phi0 alpha).

@@

Lemma phi0_le (alpha : Ord) : alpha <= phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_phi0 (alpha : Ord):
  alpha < epsilon0 -> alpha < phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Cr_fun : forall alpha : Ord,
       (forall beta : Ord, beta < alpha -> Ensemble Ord) ->
        Ensemble Ord 
:= 
   fun (alpha :Ord)
        (Cr : forall beta, 
                beta < alpha -> Ensemble Ord) 
        (x : Ord) => (
       (alpha = zero /\ AP x) \/
       (zero < alpha /\
        forall beta (H:beta < alpha),
          the_ordering_segment (Cr beta H) x /\ ord (Cr  beta H) x = x)).

@@

Definition sorted (l: cnf_t) :=
  LocallySorted (fun alpha beta => beta <= alpha) l.

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0 : 
 forall l alpha, 
   is_cnf_of alpha l ->  alpha < epsilon0 ->
   Forall (fun beta =>  beta < alpha) l.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0_iff : 
 forall l alpha, 
   is_cnf_of alpha l ->  
   (alpha < epsilon0 <->  Forall (fun beta =>  beta < alpha) l).
\end{Coqsrc}

@@

Theorem inject_mono (beta gamma : T1) :
  T1.lt  beta gamma -> 
  T1.nf beta -> T1.nf gamma -> 
  inject beta < inject gamma.

@@

\begin{Coqsrc}
Theorem inject_lt_epsilon0 (alpha : T1):
      inject alpha < epsilon0.
